Hierarchical Principal Manifold Estimation

### 2023-08-09 paper/main.tex
introduction draft

In neuroimaging settings, manifold learning algorithms are most often used to estimate the underlying manifold at the population level, with vectorized image intensities being used as the input data [Citations]. \cite{yueParameterizationWhiteMatter2016} used manifold learning to estimate a low-dimensional parameterization of the corpus callosum at the individual level. However, in many cases, the outcome of interest is framed as a comparison between different groups (e.g. treatment groups in a clinical trial).

>>>>

In neuroimaging settings, manifold learning algorithms are most often used to estimate the underlying manifold at the population level, with vectorized image intensities being used as the input data [Citations]. \cite{yueParameterizationWhiteMatter2016} used manifold learning to estimate a low-dimensional parameterization of the corpus callosum at the individual level. However, in many cases, the outcome of interest is framed as a comparison between different groups of individuals (e.g. treatment groups in a clinical trial). This goal is particularly relevant in the case of the complex neurodegenerative conditions mentioned above, where the high degree of heterogeneity in patient experiences have led to attempts to identify subgroups of patients based on a range of criteria such as symptom expression, onset, or rate of progression [Citations]. Fundamentally, if a low dimensional manifold is thought to effectively represent data related to the patient experience of a neurodegenerative condition, and the experience of that condition differs by group, then the low-dimensional representation of that experience should also differ by group.

A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations found in the dataset to which it is to be applied. To develop an estimate of the manifold underlying the hippocampi of participants in the ADNI study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants, which may belong to a latent disease subgroup for which patients tend to have more related symptom experiences or underlying physiological developments, which are in turn nested within study groups based on Alzheimer's disease status, which are finally nested within the greater study population. The goal of this article is to develop a manifold learning approach that appropriately accounts for this hierarchical structure, estimates the underlying manifold at each level of the hierarchy, and allows for statistical testing of differences between those low-dimensional manifolds.

There have been several approaches taken to address hierarchical structure in manifold learning problems. \cite{bhatiaHierarchicalManifoldLearning2012} introduced a modification of Laplacian eigenmaps to estimate manifolds from recursively subdivided regions of medical images while enforcing similar alignment of the manifolds estimated from related regions. \cite{freicheCharacterizingMyocardialIschemia2022} builds on this approach to incorporate multiple imaging modalities and automate the tuning of the algorithm's hyperparameters.

\cite{gaoHierarchicalManifoldLearning2019} proposed a two-step diffusion maps method for hierarchical data with variation over several domains, where diffusion maps is applied first to reduce the images for each individual to $d_1$ dimensions, resulting in individual-specific manifold estimates, then diffusion maps is applied again on the concatenated parameterizations to reduce over the individuals to $d_2$ dimensions, thus estimating a group-level manifold. This approach allows the first step of dimension reduction to capture the most important qualities of each image separately before reducing across subjects, but the two step procedure ensures that the manifold estimated at the individual level has no relation to the manifold estimated at the group level, posing problems for the interpretability of the embedding mappings and comparisons of manifolds between levels.

\cite{guerreroGroupconstrainedManifoldLearning2017} noted that the grouping behavior of observations recorded longitudinally from the same individual adversely impacted the creation of the neighborhood graph between data points when estimating a manifold for the entire population. The authors proposed a constraint specifying that, a given observation's collection of nearest neighbors may include at most one point from each individual or other specified grouping. This encourages a more even representation of the entire space of observations in the neighborhood graph, and may be applied to any manifold learning algorithm that features a neighborhood graph to specify proximities between observations. However, it simply mitigates the effects of the nesting structure on a population-level manifold estimate rather than enabling a detailed comparison of manifold estimates for each group.

This article builds on the principal manifold estimation (PME) algorithm introduced in \cite{mengPrincipalManifoldEstimation2021}, which uses an iterative procedure to specify an embedding mapping from the low-dimensional manifold to a high-dimensional space that takes the form of a smoothing spline. Drawing from literature in hierarchical and additive models, we replace the smoothing spline in the PME algorithm with an additive smoothing spline model (\cite{gelmanDataAnalysisUsing2007}, \cite{hastieGeneralizedAdditiveModels1990}, \cite{gelmanBayesianDataAnalysis2014}). While the specific model structure may change to meet the needs of the application, \cite{brumbackSmoothingSplineModels1998} and \cite{schulamFrameworkIndividualizingPredictions2015a} provide examples of the general model framework used. This ultimately yields a nonlinear dimension reduction algorithm that takes advantage of a statistically principled model to address nesting structures present in the input data. As it is based on the PME framework, the algorithm outputs an explicit representation of the embedding function, making it possible to both generate new data in high-dimensional space using low-dimensional parameters and map new input data to the estimated manifold. Smoothing spline analysis of variance allows the statistical comparison of the estimated manifolds between groups or individuals (\cite{wangMixedEffectsSmoothing1998}).

The following sections are organized as follows: Section 2 reviews additive models, specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper. Section 3 applies this method to several simulated datasets using a variety of underlying manifold structures and tests performance on the MNIST dataset. Section 4 demonstrates how the model specified in Section 2 may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset, while Section 5 provides a concluding discussion of the proposed algorithm and opportunities for further study.
